Detecting Motifs in Knowledge Graphs using Compression

### 2018-09-05 kgmotifs/kgmotifs.tex
deadline approaching

\documentclass[11pt]{article}

@@

\newcommand{\G}{{\cal G}}
\newcommand{\I}{{\cal I}}
\newcommand{\tab}{\hspace*{5mm}}
\newcommand{\from}{\leftarrow}

@@

\title{Detecting Motifs in Knowledge Graphs using Compression}

@@

\begin{abstract}
\noindent We introduce a method to detect \emph{network motifs} in knowledge graphs. Network motifs are useful patterns or meaningful subunits of the graph that recur frequently. We introduce a scalable approach for detecting such motifs in large knowledge graphs, inspired by recent work for simple graphs, and show that the motifs returned reflect the basic structure of the graph. Specifically, we show that common motifs reflect graph patterns used in common queries, basic schematic units of the graph and meaningful functional subunits, for various knowledge graphs.
\end{abstract}

\noindent The Linked Open Data cloud contains a wealth of knowledge graphs, and is growing quickly. At the time of writing the average knowledge graph contains over 65\,000 edges, with many datasets containing more than a million.\footnotemark~For such large graphs, it can be difficult to see the forest for the trees: how is the graph structured at the lowest level? What kind of things can I ask of what types of entities? What are small, recurring patterns that might represent a novel insight into the data?

\footnotetext{\url{http://stats.lod2.eu/stats}}

In the domain of plain (unlabelled) graphs, \emph{network motifs} \cite{milo2002network} were recently introduced as a tool to provide users of large graphs insight into their data. Network motifs are small subgraphs whose frequency is unexpected with respect to a null model. That is, for a given graph dataset $g$ and small graph $m$, we count the frequency $F(m, g)$: how often $m$ occurs in $g$ as a subgraph. We define a probability distribution over graphs $p^\text{null}(g)$, and estimate the probability that a graph sampled from $p^\text{null}$ contains more instances of $m$ than observed in our data: $p^\text{null}(F(m, G) \geq F(m, g))$, where $G$ is a random variable representing a graph. If this probability is low (commonly, below 0.05), we consider $m$ a motif. \footnotemark

Unfortunately, counting all occurrences of a motif in a network is an expensive operation. One that we need to perform not only on the data, but also on many samples from the null model, in order to estimate the probability described above.

\footnotetext{This procedure has the structure of a hypothesis test, but it is important not to interpret it as statistical evidence for the meaningfulness of the motif. The only thing it proves (in a frequentist statistical sense) is that $p^\text{null}$ is not the true source of the data. This is usually not a surprise: we are rarely able to model all aspects of a realistic data-generating process in a single distribution. The $p$-values used in motif analysis should be interpreted strictly as \emph{heuristics}. See also \cite{bloem2017large}.}

In \cite{bloem2017large}, an alternative method is presented that uses \emph{compression} as a heuristic for motif relevance: the better a motif compresses the data, the more likely it is to be meaningful. It is shown that this principle can be implemented in a similar sort of hypothesis test for a null model, allowing a comparable workflow to classical motif analysis. Crucially, this method scales to graphs with billions of edges, suggesting that it might be suitable for the scale of knowledge graphs we encounter in the LOD cloud.

In this paper, we extend the compression-based motif analysis to Knowledge Graphs. For the purposes of this research we define Knowledge graphs as labeled multigraphs. Nodes are labeled with entity names, and links are labeled with relations. We extend the definition of a motif to that of a basic graph pattern: that is, a motif is a small graph with \emph{some} of its nodes and links labeled. The motif \emph{occurs} in a graph where both the graph structure and the labels match. The unlabeled nodes and links are free to contain whatever label when the motif is mapped to the data.

To maintain the connection to graph pattern queries, we do not require the motifs to be \emph{induced} subgraphs: that is, when a motif is mapped to a graph, the graph can contain additional links that are not specified by the motif.

We perform several experiments to show that our method returns meaningful subgraphs. First, we look at several graphs for which schema information is provided. That is, we know the basic properties available for each entity. We show that the motifs found by our method correspond well to the schema, and compare performance against several basic benchmarks. Second, we run motif analysis on several datasets for which a list of real SPARQL queries entered by users is available. From these, we extract the basic graph patterns, and show that (a) these correspond to good motifs using our criterion and (b) our method can return such motifs with good precision and recall.

All code and benchmark datasets used in the paper are available, under open licenses.\footnotemark

\footnotetext{\url{...}}

% TODO: Make illustration

@@

\paragraph{Preliminaries}

% TODO: General purpose codes: natural numbers, integers, sequences.

@@

We will start with some formal definitions of our basic ingredients. To comfortably define probabilities and codes on graphs, we will slightly deviate from conventional definitions. Specifically, we will analyze the \emph{structure} of knowledge graphs only, ignoring any meaning they have outside the graph. For instance, we will distinguish between links with different labels, but we will ignore any internal structure of the label itself (such as shared IRI prefixes). Under this assumption we can assume that the node and link labels of our knowledge graphs are simply natural numbers. This leads to the following definition.

A \emph{knowledge graph} $G$, hereafter known simply as a \emph{graph} is a, is tuple $G = (v_G, r_G, E_G)$. $v_G \in \N$ is the number of nodes in the graph, and $r_G$ is the number of relations. We will define the nodeset as $V_G = \{0, \ldots, v_G-1\}$ and the relation-set as $R_G = \{0, \ldots, r_G\}$. The \emph{tripleset} $E_G \subset V_G \times R \times V_G$ defines the links of the graph and their labels. \footnotemark

\footnotetext{Commonly, the elements of the nodeset of a graph are defined to be simple atoms from some larger unspecified universe. However, since we are discussing sampling graphs, we need to specify where these atoms come from. This definition specifies that graphs are essentially unlabeled, but nodes do have a definite ordering (i.e. two graphs can be isomorphic without being the same graph).}

As shown in Figure~\ref{figure:diagram}, the patterns that we aim to find are partially labeled: some nodes and edges are labeled with negative integers. In other words, the motif is a \emph{basic graph pattern}, and the edges and nodes marked with negative integers are variables. We will define a pattern in the same way we define a graph, but we will extend its nodeset en relationset. A pattern $M$ for graph $G$ is a tuple $(v'_M, r'_M, G, E_M)$. $v'_M$ and $r'_M$ indicate the number of variable nodes and links respectively. $M$'s nodeset is defined as $V_M = \{-v'_M, \ldots, v_G-1\}$ and the relationset as $R_M = \{-r'_M, \ldots, r_G-1\}$, with $E_M \subset V_M \times R_M \times V_M$, as for graphs. In short, nodes can be labeled with nonnegative integers referring to $G$'s nodes or with negative integers representing a variable node, and similar for relations.

An \emph{instance} for $M$ in $G$ is a pair of sequences of integers: $I = (I^n, I^r)$. $I^n$ is a sequence of distinct integers of length $v'_M$. $I^r$ is a sequence of non-distinct integers of length $r'_M$. For each link $(s, p, o) \in E_M$ containing a negative $s$, $p$ or $o$, there is a corresponding link in $E_G$ with a negative $s$ replaced by $I^n_{-s}$, a negative $o$ replaced by $I^n_{-o}$, and a negative $p$ replaced by $I^r_{-p}$. Put simply: for a pattern to match, variable links marked with the same negative integer, must map to the same relation in order for the pattern to match, but variable links labeled with different negative integers \emph{may} map to the same relation. Variable nodes are always labeled distinctly and may never map to the same node in $G$. An instance describes a subgraph of $G$ that \emph{matches} the pattern $M$.

We will first assume that a target pattern $M$ is given for the data $G$ and that we have a set of instances $\I$. Moreover, we require that all instances in $\I$ are mutually disjoint: no two subgraphs defined by member of $\I$ may share an edge, but nodes may be shared. In most settings we do not actually have a given $M$ and $\I$ and we will need to search for them. We describe a simple and fast search algorithm in Section~\ref{section:search}.

\subsection{Relevance test}

@@

We will first describe our method under the assumption that a good null model $p^\text{null}$ is available, and that $- \log p^\text{null}(G)$ can be efficiently computed. In Section~\ref{section:null-model}, we explain which null model we use for our experiments, and how it is implemented.

In the design of our method, we will constantly aim to find a trade-off between completeness and efficiency that allows the method to scale to very large graphs. Specifically, when we economize, we will only do so in a way that makes the hypothesis test described in Section~\ref{section:relevance-test} \emph{more conservative}.

@@

As described above, we can perform our relevance test with any compression method which exploits the pattern $M$, and its instances $I$ to store the graph efficiently. The better our method, the more motifs we will find. Note that there is no need for our code to be optimal in any sense. So long as we accept that we won't find all motifs that exist, we are free to trade off compression performance against efficiency of computation.

We store the graph by encoding various aspects, one after the other. The information in all of these together is sufficient to reconstruct the graph. Note that everything is stored using prefix-free codes, so that we can simply sum the codelengths we get for each aspect to get the codelength for the complete graph.

We also assume that we are given a code $L^\text{base}$ for generic knowledge graphs.

@@

\begin{description}
 \item[the graph dimensions] We first store $|V_G|$, $|E_G|$ and $|R_G|$ using the generic code $L^{\N}(\cdot)$. We also store the size of $\I$.
 \item[the pattern] Let $M'$ be the unlabeled graph representation of $M$: that is, $M'$ is a plain graph (not a pattern), $V_{M'}$ contains only the first $V_{M'}$ natural numbers and similarly for $R_{M'}$. Otherwise, $M$ is isomorphic to $M'$: the node and edge labels of $M$ can be mapped to those of $M'$ so that $M$ only contains an edge $E$ if and only if $M'$ contains the edge resulting from mapping $E$'s symbols. 
 \item[the template] This is the graph, minus all links occurring in  instances. Let $E_G'$ be $E_G$ minus any link occurring in any member of $\cal I$. We then store $(V_G, E_G')$ using $L^\text{base}(\cdot)$.
 \item[the instances] In order to reconstruct the graph from the template, we need to know which nodes correspond to instances of the motif, and in what order. For each node in $M'$, we create a list of length $\cal |I|$ of the nodes in $G$ to which it maps for each instance. For the non-variable nodes of the pattern, this will be a list simply repeating one node node of $G$. For the variable nodes, it will be a list containing some subset of the nodes in $G$.
 We store each list using the DM code. It may seem wasteful to store a list repeating a single element (instead of storing a labeled version of the pattern), but under the DM code such lists will be stored efficiently, using only a small, constant number of bits. The more varied the list, the more bits will be required to store it.
 \item[the edge labels] We now have enough information to reconstruct the graph structure. All that remains to fully reconstruct $G$ is to label the links of the motif instances. As for the nodes, we create one list of length $\cal I$ for each type of link in $M'$, which we store using the DM code.
\end{description}

@@

\begin{pseudo}[tb]
\caption{The motif code $L^\text{motif}(G ; M, {\cal I}, L^\text{base})$. Note that the nodes and relations of the graph are integers.}
\label{algorithm:motif-code}
{ 
\textbf{function} $\text{codelength}(G; M, {\cal I}, L^\text{base})$:\\
\tab\tab a graph $G$, a pattern $M$\\ 
\tab\tab a list $\cal I$ of instances of $M$ in $G$, a code $L^\text{base}$ on knowledge graphs.\\
\\
\tab $b_\text{dim} \leftarrow L^\N(|V_G|) + L^\N(|R_G|) + L^\N(|E_G|) + L^\N(|\I|)$ \\

\tab \emph{// Turn the pattern into a normal knowledge graph}\\
\tab $E_{M'} = \left\{(\text{i}(s, V_M),  \text{i}(p, R_M), \text{i}(o, V_M)) \mid (s, p, o) \in E_M\right\}$\\
\tab $M' = ([0 ..  |V_M|-1], [0 .. |R_M|-1], E_{M'})$\\
\tab $b_\text{pattern} \leftarrow L^\text{base}(M')$ \\

\tab $E'_G = E_G - \cup_{\I \in {\cal I}} I$ \hfill \emph{// Store the graph, with instances removed}\\
\tab $b_\text{template} = L_\text{base}((V_G, R_G, E'_G))$\\

\tab $b_\text{labels} \leftarrow 0$ \hfill \emph{// Store the node labels} \\
\tab \textbf{for} $n$ in $V_M$\\
\tab\tab $\I_n = \langle v(n, I_1), v(n, I_2), ... \rangle$\\
\tab\tab\tab with $v(n, I)$ the label (from $V_G$) of node $n$ in instance $I$\\
\tab\tab $b_\text{labels} \leftarrow b_\text{labels} + L^{DM}(\I_n)$\\

>>>>

\documentclass[letterpaper]{article} %DO NOT CHANGE THIS

\usepackage{aaai19}    % Required
\usepackage{times}     % Required
\usepackage{helvet}    % Required
\usepackage{courier}   % Required
\usepackage{url}       % Required
\usepackage{graphicx}  % Required
\frenchspacing         % Required
\setlength{\pdfpagewidth}{8.5in}  % Required
\setlength{\pdfpageheight}{11in}  % Required

% PDF Info Is Required:
\pdfinfo{
/Title (Detecting Motifs in Knowledge Graphs using Compression)
/Author (...)}

\setcounter{secnumdepth}{0}

@@

\newcommand{\X}{{\cal X}}
\newcommand{\B}{{\mathbb B}}
\newcommand{\G}{{\cal G}}
\newcommand{\cS}{{\cal S}}

\newcommand{\I}{{\cal I}}
\newcommand{\tab}{\hspace*{5mm}}
\newcommand{\from}{\leftarrow}

@@

\title{Detecting Motifs in Knowledge Graphs using Compression}
\author{
Author names withheld
% Peter Bloem\\
% Knowledge Representation \& Reasoning Group\\
% VU University\\
% vu@peterbloem.nl\\
}

@@

\begin{abstract}
\noindent We introduce a method to detect \emph{network motifs} in knowledge graphs. Network motifs are useful patterns or meaningful subunits of the graph that recur frequently. We extend the common definition of a network motif to coincide with a \emph{basic graph pattern}, commonly used in the knowledge graph literature. We introduce an approach, inspired by recent work for simple graphs, for inducing these for a given knowledge knowledge graph, and show that the motifs returned reflect the basic structure of the graph. Specifically we show that in random graphs, no motifs are found, and that when we insert a motif artificially, it can be detected, even if as few as 10 instances are inserted. Finally, we show the results of motif induction on three real-world knowledge graphs, and show that our method provides good suggestions for motifs that reflect the structure of the graph in an intuitive way.
\end{abstract}

\noindent \emph{Knowledge graphs} are an extremely versatile way of storing knowledge. They allow knowledge to be encoded without a predefined schema, they allow different amounts of data to be encoded for each known entity and they allow different modalities to be integrated seamlessly \cite{wilcke2017knowledge}. This versatility comes at a price. For a given knowledge graph, it can be difficult to see the forest for the trees: how is the graph structured at the lowest level? What kind of things can I ask of what types of entities? What are small, recurring patterns that might represent a novel insight into the data?

In the domain of unlabeled simple graphs, \emph{network motifs} \cite{milo2002network} were introduced as a tool to provide insight into local graph structure. Network motifs are small subgraphs whose frequency in the graph is unexpected with respect to a \emph{null model}. That is, for a given graph dataset $G$ and small graph $M$, we count the frequency $f(M, G)$: how often $M$ occurs in $G$ as a subgraph. We define a probability distribution over graphs $p^\text{null}(\G)$, and estimate the probability that a graph sampled from $p^\text{null}$ contains more instances of $M$ than observed in our data: $p^\text{null}(f(M, \G) \geq f(M, G))$, where $\G$ is a random variable representing a graph. If this probability is very low (commonly, below 0.05), we consider $m$ a motif. \footnotemark

\begin{figure*}[bth]
  \centering
    \includegraphics[width=\textwidth]{example.pdf}
    \caption{An example of the principle behind our motif code. a) A basic knowledge graph. Note that in our analysis, we consider only the indices of the nodes and relations, not their labels. b) A motif that occurs frequently. c) A compressed representation; we remove all edges that are part of an occurrence of the motif. We then store separately which nodes match the motif. The motif, together with the template graph and the matches can be used to reconstruct the graph. We take this as a heuristic: the better the motif compresses, the more characteristic it is likely to be for the graph.}
    \label{figure:example}
  \label{figure:codes}
\end{figure*}

Unfortunately, estimating this probability usually requires repeating the subgraph count on many samples from the null model. To avoid this costly operation, \cite{bloem2017large} introduces  an alternative method, using \emph{compression} as a heuristic for motif relevance: the better a motif compresses the data, the more likely it is to be meaningful. It is shown that this principle can be implemented in a similar sort of hypothesis test for a null model, allowing a comparable workflow to classical motif analysis.

\footnotetext{This procedure has the structure of a hypothesis test, but it is important not to interpret it as \emph{statistical evidence} for the meaningfulness of the motif. The only thing it shows (in a frequentist statistical sense) is that $p^\text{null}$ is not the true source of the data. This is usually not a surprise: we are rarely able to model all aspects of a realistic data-generating process in a single distribution. The $p$-values used in motif analysis should be interpreted strictly as \emph{heuristics}. See also \cite{bloem2017large}.}

In this paper, we extend the compression-based motif analysis to knowledge graphs. For the purposes of this research we define knowledge graphs as labeled, directed multigraphs. Nodes are uniquely labeled with entity names, and links are non-uniquely labeled with relations.  We extend the definition of a motif to that of a \emph{basic graph pattern}: that is, a motif is a small graph with \emph{some} of its nodes and links labeled with entities and relations as they occur in the graph. Other nodes and links are labeled as \emph{variables}. A pattern matches at a particular location in the graph, if the variables can be replaced with values from the graph, and the motif becomes a subgraph as a result. 
The intuition behind our method is that the more compactly a pattern compresses the graph, the better the motif. Figure~\ref{figure:example} shows an illustration of the basic principle. In the preliminaries section below, we justify this intuition more formally.

We perform several experiments to show that our method returns meaningful subgraphs. First we test the intuition that a random graph shoudl contain no motifs. We also show that when we artifically insert motifs into a random graph, we can then detect these as motifs, sometimes with as few as 10 instances inserted into the graph. Finally, we show the results of motif analysis on three real-world knowledge graphs.

All code and datasets used in this paper are available, under open licenses.\footnotemark

\footnotetext{URL withheld for purposes of blind reviewing.}

@@

Network motifs for unlabeled simple graphs were introduced in \cite{milo2002network}. A more comprehensive overview of the related literature can be found in \cite[Section~1.1]{bloem2017large}. In \cite{bloem2017large}, the principle of Minimum Description Length (MDL) was first connected to motif analysis. However, the idea had earlier been exploited for detecting meaningful subgraphs in the SUBDUE algorithm \cite{cook1994substructure}.

A few other methods have been proposed for inducing the structure of a given knowledge graph. In \cite{pham2015deriving}, the authors use the principle of characteristic sets to characterize a knowledge graph in terms of the star patterns it contains. In \cite{pham2016exploiting}, they show that the majority of the LOD cloud can be efficiently described using such principles, showing the highly tabular structure of many knowledge graphs. In \cite{volker2011statistical}, association rules mining is used to induce basic patterns in the graph.

To the best of our knowledge, ours is the first method presented that can potentially induce any basic graph pattern.

\subsection{Preliminaries}

\paragraph{Minimum Description Length}
Our method is based on the MDL principle: that we should favour models that compress the data. We will show briefly how this intuition can be made mathematically precise. For more details, we refer the reader to \cite{grunwald2007minimum} for MDL in general, and to \cite{bloem2017large}, for a more extensive discussion these principles in the domain of graph analysis.

Let $\B$ be the set of all finite-length binary strings. We use $|b|$ to represent the length of $b \in \B$. Let $\log(x) = \log_2(x)$. A \emph{code} for a set of objects $\cal X$ (such as graphs) is an injective function $f: {\cal X} \to \B$, mapping objects to binary code words. All codes in this paper are \emph{prefix-free}: no code word is the prefix of another. We will denote a \emph{codelength function} with the letter $L$, ie. $L(x) = |f(x)|$. We commonly compute $L(x)$ directly, without first computing $f(x)$ and we refer to $L(x)$ as a code.

A well known result in information theory is the association between codes and probability distributions, implied by the \emph{Kraft inequality}: for each probability distribution $p$ on $\cal X$, there exists a prefix-free code $L$ such that for all $x \in \cal X$: $- \log p(x) \leq L(x) < -\log p(x) + 1$. Inversely, for every prefix-free code $L$ for $\cal X$, there exists a probability distribution $p$ such that for all $x \in \cal X$: $p(x) = 2^{-L(x)}$. For proofs, see \cite[Section~3.2.1]{grunwald2007minimum} or \cite[Theorem~5.2.1]{cover2006elements}. To explain the intuition, note that we can easily transform a code $L$ into a sampling algorithm for $p$ by feeding the decoding function random bits until it produces an output. To transform a probability distribution to a code, techniques like arithmetic coding \cite{rissanen1979arithmetic} can be used.\footnotemark 
\footnotetext{
As explained in \cite[page 96]{grunwald2007minimum}, the fact that $-\log p^*(x)$ is real-valued and $L^*(x)$ is integer-valued can be safely ignored and we may \emph{identify} codes with probability distributions, allowing codes to take non-integer values. 
}

When we need to encode a single choice from a finite set $S$ of options, we will often use the code with length $\log |S|$, corresponding to a uniform distribution on $S$. In some cases, we can allow codes with multiple codewords for a single object. If a code $L$ has multiple codewords for some object  $x$, we may indicate the choice for a particular codeword by a parameter $a$ as $L(x; a)$.

\paragraph{Relevance testing} We will use the MDL principle to perform a hypothesis test. Assume we have some data $x$ and a null hypothesis that it was sampled from distribution $p^\text{null}$. Let $L^\text{null}$ be the corresponding codelength. A simple but crucial result, known as the \emph{no-hypercompression inequality} \cite[p103]{grunwald2007minimum} tells us that the probability of sampling data $x$ from $p^\text{null}$ that can be described shorter than $L^\text{null}(x)$ by $k$ or more bits, \emph{using any code} is less than $2^{-k}$.
Thus, a standard MDL hypothesis test would consist of stating the null hypothesis in terms of a code $L^\text{null}$, designing a code (before seeing the data) which compresses the data better than $L^\text{null}$ by, say, 10 bits and rejecting the null hypothesis with confidence $2^{-10}$. For a longer, more intuitive explanation of this principle in pattern induction, we refer the reader to \cite[Section~2]{bloem2017large}.

We must note again, that when we use this procedure to find motifs, we are not providing evidence for the hypothesis that the motif is ``correct.'' The only thing we are statistically proving is that the null-model is incorrect. Nevertheless, hypothesis testing has shown to be a valuable heuristic in motif analysis.

\paragraph{Common codes}
In the construction of our codes for graphs, we require some simpler codes as building blocks. First, when we store any positive integer $n$, we do so with the code corresponding to the distribution $p^\N(n) = 1/(n(n+1))$, and denote it $L^\N(n)$. For nonnegative numbers we add 1 to the argument. for the full range of integers, we add an extra bit for the sign, and then use the first code for negative integers and the second for positive ones.

We will often need to encode sequences of integers. These will be highly skewed, with only a subset of integers occurring frequently, and others occurring infrequently or not at all. As noted in \cite{de2016names} a code based on the Pitman-Yor model \cite{pitman1997two} is very effective in such situations. Let $S = \langle S_1, ..., S_n\rangle$ be a sequence of integers of length $n$. We first store its members $m(s)$ in the order in which they first occur: we first store $n$ and the first member using the codes described above. The rest of the members are stored sequentially using the distance from the current to the next member.

Let $f(A, B)$ be the frequency of symbol $A$ in sequence $B$.We then store the complete sequence using the code corresponding to the following distribution:
\begin{align*}
&p^\text{PY}_{\alpha,d}(S) = \prod_{i \in [1,k]} \text{PY}_{\alpha,d}(S_i\mid S_{1:i-1}) \\
&\text{PY}_{\alpha,d}(S_i \mid S') = 
\begin{cases}
\frac{\alpha - d |m(S)|}{|m(S)| + \alpha} & \text{if} f(S_i, S') = 0 \\ 
\frac{f(S_i, S') - d}{|m(S)| + \alpha} & \text{otherwise}
\end{cases}
\end{align*}
See \cite{de2016names} for a more intuitive explanation. In all experiments we use $\alpha = 0.5$, $d=0.1$. We will refer to the total for a sequence  codelength under this procedure as $L^PY(S)$

@@

We will start with some formal definitions of our basic ingredients. To comfortably define probabilities and codes on graphs, we will slightly deviate from conventional definitions. Specifically, we will analyse the \emph{structure} of knowledge graphs only, ignoring any meaning they have outside the graph. For instance, we will distinguish between links with different labels, but we will ignore any internal structure of the label itself (such as shared IRI prefixes). Under this assumption we can assume that the node and link labels of our knowledge graphs are simply natural numbers.\footnotemark This leads to the following definition.

\footnotetext{For practitioners this restriction is not noticeable, as the indices can simply be mapped back to the original strings when the found motifs are presented.}

A \emph{knowledge graph} $G$, hereafter simply a \emph{graph}, is a tuple $G = (v_G, r_G, E_G)$. $v_G \in \N$ is the number of nodes in the graph, and $r_G \in \N$ is the number of relations. We define the nodeset as $V_G = \{0, \ldots, v_G-1\}$ and the relation-set as $R_G = \{0, \ldots, r_G\}$. The \emph{tripleset} $E_G \subset V_G \times R \times V_G$ determines the edges of the graph and their labels.

As shown in Figure~\ref{figure:example}, the patterns that we aim to find are partially labeled: some nodes and edges are labeled with as variables. We will represent these with negative integers. A pattern $M$ for graph $G$ is a tuple $(V_M, R_M, G, E_M)$. Let $v_M$ and $r_M$ indicate the number of variable nodes and variable links in $M$ respectively, then $V_M \subseteq \{-v'_M, \ldots, v_G-1\}$ and $R_M \subseteq \{-(r'_M+v'M), \ldots,-v'M, 0,\ldots, r_G-1\}$, with $E_M \subset V_M \times R_M \times V_M$, as for regular graphs. That is; nodes can be labeled either with nonnegative integers referring to $G$'s nodes or with negative integers representing a variable node, and similar for relations. The negative integers are always contiguous within a single graph, with the highest representing the node labels and the lowest representing the edge labels.

An \emph{instance} for $M$ in $G$ is a pair of sequences of integers: $I = (I^n, I^r)$. $I^n$ is a sequence of distinct integers of length $v'_M$. $I^r$ is a sequence of non-distinct integers of length $r'_M$. For each edge $(s, p, o) \in E_M$ containing a negative $s$, $p$ or $o$, there is a corresponding link in $E_G$ with a negative $s$ replaced by $I^n_{-s}$, a negative $o$ replaced by $I^n_{-o}$, and a negative $p$ replaced by $I^r_{-p - v'_M}$. Put simply: for a pattern to match, variable edges marked with the same negative integer, must map to the same relation in order for the pattern to match, but variable links labeled with different negative integers \emph{may} map to the same relation. Variable nodes are always labeled distinctly and may never map to the same node in $G$. An instance describes a subgraph of $G$ that \emph{matches} the pattern $M$. Each edge in the motif may only match one edge in the graph. In other words, the occurrence of the motif in the graph must have as many edges as the motif itself.

We will first assume that a target pattern $M$ is given for the data $G$ and that we have a set of instances $\I$. Moreover, we require that all instances in $\I$ are mutually disjoint: no two subgraphs defined by member of $\I$ may share an edge, but nodes may be shared. In most settings we do not actually have a given $M$ and $\I$ and we will need to search for them. We describe a simple and search algorithm in Section~\ref{section:search}.

@@

The most common null model for classical motif analysis is the degree-sequence model (also known a the configuration model \cite{newman2010networks}): a uniform distribution over all graphs with a particular degree sequence. We extend this to knowledge graphs by also including the degree of each relation: that is, the frequency with which it occurs in the tripleset. Let a \emph{degree sequence} $D$ of length $n$ be a triple of three integer sequences: $(D^\text{in}, D^\text{rel}, D^\text{out})$. If $D$ is the degree sequence of a graph, then node $i$ has $D^\text{in}_i$ incoming links,  $D^\text{out}_i$ outgoing links and for each relation $r$, there are $D^\text{rel}_r$ triples.

Let $\G_D$ be the set of all graphs with degree sequence $D$. Then the configuration model can be expressed simply as
\[
p^C(G) = \frac{1}{|\G_D|}
\]
for any $G$ that satisfies $D$ and $p(G) = 0$ otherwise. Unfortunately, there is no efficient way to compute $|\G_D|$ and even approximations tend to be costly for large graphs. Following the approach in \cite{bloem2017large}, we define a fast approximation to the configuration model, which works well in practice for motif detection.

We can describe a knowledge graph by three length-$m$ integer sequences: $S$, $P$, $O$, such that $\{(S_j, P_j, O_j)\}_j$ is the graph's tripleset. If the graph satisfies degree sequence $D$, then we know that $S$ should contain node $j$ $D^\text{out}_j$ times, $P$ should contain relation $r$ $D^\text{rel}_r$ times and $O$ should contain node $j$ $D^\text{in}_j$ times.  Let ${\cal S}_D$ be the set of all such triples of integer sequences satisfying $D$. We have 
\[
|{\cal S}_D| =
 {m \choose {D_1^\text{out}, \ldots, D_n^\text{out}} }
 {m \choose {D_1^\text{rel}, \ldots, D_{|R_G|}^\text{rel}} }
 {m \choose {D_1^\text{in}, \ldots, D_n^\text{in}} } \text{.}
\]

While every member of ${\cal S}_D$ represents a valid graph satisfying $D$, many graphs are represented multiple times. Firstly, many elements of ${S}_D$ contain the same link multiple times. We call the set without these elements ${\cal S'}_D \subset {\cal S}_D$. Secondly the links of the graph are listed in arbitrary order; if we apply the same permutation to all three lists $S$, $P$ and $O$, we get a new representation of the same graph. Since we know that any element in ${\cal S}'_D$ contains only unique triples, we know that each graph is present exactly $m!$ times. This gives us
\[
|\G_D| = |{\cal S}'_D| \frac{1}{m!} \leq  |{\cal S}_D| \frac{1}{m!} \text{.}
\]

We can thus use 
\[
p^\text{EL}_D(G) =  \frac{m!}{{m \choose {D_1^\text{out}, \ldots, D_n^\text{out}} }
 {m \choose {D_1^\text{rel}, \ldots, D_{|R_G|}^\text{rel}} }
 {m \choose {D_1^\text{in}, \ldots, D_n^\text{in}} }} \leq p^C(G) \text{.}
\]

Filling in the definition of the multinomial coefficient, and rewriting, we get a codelength of:
\begin{align*}
- &\log p^\text{EL}_D(G) = 2 \log(m!) \\
&- \sum_i \log(D_i^\text{in}!)- \sum_i \log(D_i^\text{rel}!)- \sum_i \log(D_i^\text{out}!)
\end{align*}

as an approximation for the DS model. We call this the edgelist (EL) model. It gives a probability that always lower-bounds the configuration model, since it affords some probability mass to graphs that cannot exist.\footnotemark ~Experiments in the classical motif setting have shown that the EL model is an acceptable proxy for the DS model \cite{bloem2017large}, especially considering the extra scalability it affords.

\footnotetext{Note that we cannot simply think of $p^\text{EL}$ as a uniform model for graphs containing multiple links, since we can only divide by $m!$ if we know that all triples are unique.}

 \paragraph{Encoding D} In order to encode a graph with $L^\text{EL}_D$, we must first encode $D$. \footnote{Or, equivalently, to make $p^\text{EL}$ a complete distribution on all graphs, we must provide it with a prior on $D$.} For each of the three sequences in $D$ we use the following model:
\begin{align*}
 p(D) &= \prod_i q^{\N}(D_i) \\
 L(D) &= - \sum_i \log q^{\N}(D_i)
\end{align*}

where $1^{\N}$ is any distribution on the natural numbers. This is an optimal encoding for $D$ assuming that its members are independently drawn from $q^{\N}$. When we use $p^\text{EL}$ as the null model, we use the data distribution for $q^{\N}$ to ensure that we have a lower bound to the optimal code-lengthn (in essence, we cheat, giving the null model a slightly lower than optimal codelength). When we use $p^\text{EL}$ as part of the motif code, we must use a fair encoding, so we use the Pitman-Yor code to store each sequence in $D$.

In the design of our method, we will constantly aim to find a trade-off between completeness and efficiency that allows the method to scale to very large graphs. Specifically, when we economize, we will only do so in a way that makes the hypothesis test \emph{more conservative}.

@@

As described above, we can perform our relevance test with any compression method which exploits the pattern $M$, and its instances $\I$ to store the graph efficiently. The better our method, the more motifs we will find. Note that there is no need for our code to be optimal in any sense. So long as we accept that we won't find all motifs that exist, we are free to trade off compression performance against efficiency of computation.

We store the graph by encoding various aspects, one after the other. The information in all of these together is sufficient to reconstruct the graph. Note that everything is stored using prefix-free codes, so that we can simply concatenate the codewords (or rather, sum the codelengths we get for each aspect).

We also assume that we are given a code $L^\text{base}$ for generic knowledge graphs (in practice, this will be the null model, although the motif code is valid for any base code).

@@

\begin{description}
 \item[the graph dimensions] We first store $|V_G|$, $|E_G|$ and $|R_G|$ using the generic code $L^{\N}(\cdot)$. 
 \item[the pattern] We store the structure of the pattern using the base code, and its labels as a sequence using the Pitman-Yor code.
 \item[the template] This is the graph, minus all links occurring in instances of $M$. Let $E_G'$ be $E_G$ minus any link occurring in any member of $\cal I$. We then store $(V_G, E_G')$ using $L^\text{base}(\cdot)$.
 \item[the instances] To store the instances, we view the connections between the nodes made by motifs as a hypergraph, and we extend the EL code to store it. The details are given below.
\end{description}

@@

\begin{pseudo}[tb]
\caption{The motif code $L^\text{motif}(G ; M, {\cal I}, L^\text{base})$. Note that the nodes and relations of the graph are integers.}
\label{algorithm:motif-code}
{ 
\textbf{function} $\text{codelength}(G; M, {\cal I}, L^\text{base})$:\\
\tab\tab a graph $G$, a pattern $M$\\ 
\tab\tab instances $\cal I$ of $M$ in $G$, a code $L^\text{base}$.\\
\\
$b_\text{dim} \leftarrow L^\N(|V_G|) + L^\N(|R_G|) + L^\N(|E_G|)$ \\

\emph{\# Turn the pattern into a normal knowledge graph}\\
$E_{M'} \leftarrow$ the edges of $M$ with contiguous integer labels \\
$M' \leftarrow (|V_M|, |R_M|, E_{M'})$ \\
$S_M \leftarrow$ the labels of M in canonical order \\
$b_\text{pattern} \leftarrow L^\text{base}(M') + L^{PY}(S_M) $\\

\emph{\# Store the template graph}\\
$E'_G \leftarrow E_G - \cup_{\I \in {\cal I}} \text{triples}(I)$ \\
$b_\text{template} \leftarrow L_\text{base}((V_G, R_G, E'_G))$\\

$b_\text{instances} \leftarrow -\log p_M(\I) + \sum_{D \in D^{\I}} L^{PY}(D)$\\

\textbf{return} $b_\text{dim} + b_\text{pattern} + b_\text{template} + b_\text{instances}$\\
}
\end{pseudo}

\subsection{Encoding motif instances}

To encode a list of instances $\I$ of a given pattern $M$, we generalize  the idea of the edgelist model described above. To understand the following ,it may be instructive to think of a single triple as an instance for the pattern
\[
\texttt{?n1 ?rel ?n2}.
\]
 We can now think of the edgelist model as storing a sequence of instances for this pattern, which together make up the entire knowledge graph.

 To generalize this notion to arbitrary patterns, to be defined for a given template graph, we define the \emph{degree constraint} $D^\I$ of a list of instamces for a given pattern as follows: for each variable node $i$ in the pattern, the degree sequence provides an integer sequence $D^i$ of length $v_G$, indicating how often each node in the completed knowledge graph takes that position in the pattern. Similarly, for each variable link $j$ in the pattern, the degree sequence provides an integer sequence $C^j$ of length $r_G$ indicating for each relation how often it takes that position in the pattern.

 We store these sequences in the same manner as the degree sequence of the template graph, using the Pitman-Yor code for each.
